Species sympatry shapes brain size evolution in Primates

### 2022-01-28 Article/Article_forsubmission.Rmd
Update for submission - Separation Sup mat and Main text - Add Rmd word compilation

#I follow: https://stackoverflow.com/questions/52918716/authors-and-affiliations-in-the-yaml-of-rmarkdown for author display
author: 
  - name: Benjamin Robira
    email: benjamin.robira@normalesup.org
    institute: [CEFE, MH]
    correspondence: true
  - name: Benoît Perez-Lamarque
    email: benoit.perez@ens.psl.eu
    institute: [ENS, MNHN]
institute:
  - CEFE: Centre d’Écologie Fonctionnelle et Évolutive, Université de Montpellier & CNRS, Montpellier, France.
  - MH: Anthropologie et Ethnobiologie, Centre National de la Recherche Scientifique/Muséum National d'Histoire Naturelle, University Paris Diderot, Sorbonne Paris Cité, Musée de l'Homme, Paris, France.
  - ENS: Institut de Biologie de l’École Normale Supérieure (IBENS), École Normale Supérieure, CNRS, INSERM, Université PSL, Paris, France.
  - MNHN: Institut de Systématique, Évolution, Biodiversité (ISYEB), Muséum National d’Histoire Naturelle, CNRS, Sorbonne Université, EPHE, Université des Antilles, Paris, France.

@@

output: 
 #      - '--lua-filter=scholarly-metadata.lua'
 #      - '--lua-filter=author-info-blocks.lua'
 bookdown::pdf_book:
 #bookdown::word_document2:
   number_sections: false
   toc: false
   citation_package: biblatex
   latex_engine: pdflatex
   fig_caption: true
   pandoc_args:
     - '--lua-filter=scholarly-metadata.lua'
     - '--lua-filter=author-info-blocks.lua'
always_allow_html: true

bibliography: bibliographyarticlepackage.bib
biblio-style: nature
biblatexoptions: [natbib=true]

@@

header-includes:
  - \usepackage{lettrine}
  # - \usepackage[nolists, nomarkers,tablesfirst]{endfloat} # For figures and tables at end
  - \usepackage{lineno} # For line numbering
  - \linenumbers # For line numbering
  - \usepackage{setspace}\doublespacing
  - \usepackage{fontawesome} #for fa symbols
  - \usepackage{tcolorbox}
  - \pagenumbering{gobble} #for no page numbering
  - \setlength{\parskip}{0em} #to remove line gaps
  - \DeclareUnicodeCharacter{2212}{-}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \newcommand{\beginsupplement}{
      \setcounter{table}{0}  
      \renewcommand{\thetable}{S\arabic{table}} 
      \setcounter{figure}{0} 
      \renewcommand{\thefigure}{S\arabic{figure}}
    } 
  - \usepackage{titling}
  - \pretitle{
    \begin{flushleft}
    \rule[-0.15in]{0.25\linewidth}{0.8ex}
    \vspace{-0.8ex}
    \hrule
    \vspace{0.3in}
    \begin{LARGE}
    \noindent
    \textbf
    }
  - \posttitle{
    \end{LARGE}\newline
    \rule[-0.15in]{0.25\linewidth}{0.8ex}
    \hrule
    \end{flushleft}
    \vspace{0.2in}}
---

@@

```{r setup, include=FALSE, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(include=FALSE, echo=FALSE, message=FALSE)
knitr::opts_chunk$set(dpi=300) # Figure resolution and size out.width = '100%', 
knitr::opts_chunk$set(fig.pos = 'p'#, fig.align = 'center'
                      ) # Places figures on pages separate from text, centered
knitr::opts_chunk$set(fig.env="figure") # Latex figure environment
knitr::opts_knit$set(eval.after = "fig.cap") #To insert R code into R figure caption
```

@@

```{r wordCount}
words <- RmdWords("Article.Rmd")
```

@@

**Abstract |** The main hypotheses related to animal intelligence evolution highlight the role of conspecifics. Yet, space is often simultaneously occupied by species sharing the same ecological guild. These sympatric heterospecifics can compete for food, thereby stimulating or hampering cognition. Considering brain size as a proxy for cognition, we used primates to test for the intertwining between species sympatry and cognition. We retraced the evolutionary history of several brain areas with evolutionary models considering or not sympatry. Sympatry-related models best predicted the evolution of brain areas related to long-term memory of interactions with the social or ecological environment, with a decrease of their size the higher the sympatry. By contrast, the whole brain or brain areas used in immediate information processing were best described by models not considering sympatry. Moreover, sympatry negatively affected primate diversification. Overall, this comparative study suggests that species sympatry contributes to shaping primate cognition and diversification. We speculate that this is due to an over-complexification of resource spatio-temporality.

@@

<!-- Reviewer possibles: Decasien, Powell, Barton, Kamilar, Harmon, Drury -->

\faTags \hspace{0.01in} **Keywords:** Brain size - Cognition - Competition - Co-occurrence - Diversification - Frugivory - Primates - Sympatry

\faInfoCircle  \hspace{0.01in} **Word Count:** `r words$num_words` \newline
<!-- Character Count: `r words$num_char` -->

@@

\initial{\textcolor{black}{C}}ognition evolution is shaped by the balance between socio-ecological drivers promoting cognitive abilities [@gonzalez2018inference] and physiological and energetic constraints limiting them [@navarrete2011energetics]. Primates are pivotal species for cognitive studies [@byrne2000evolution] because their cognition is thought to be promoted by interactions of individuals with conspecifics within the social unit [@byrne1994machiavellian; @dunbar2017there], among generations [@wilson1991molecular; @whiten2007evolution; @reader2002social; @herrmann2007humans; @tomasello2009cultural; @van2011social], between social units [@ashton2020interactions], or with the rest of their environment [@clutton1980primates; @milton1981distribution; @rosati2017foraging]. However, space is often occupied by many species belonging to the same ecological guild. Because of competition for food, we can predict that indirect interactions with sympatric heterospecifics are also likely to strongly shape the evolution of cognition, because they either affect the resource landscape by depleting it (Hypothesis 1), or the landscape of usable “social” cues to locate available food ([@Avargues-Weber2013]; Hypothesis 2). In addition, direct interactions with heterospecifics, as much as with conspecifics [@dunbar2017there], might also affect cognition (Hypothesis 3).

|   Retracing the evolutionary history of cognitive abilities proves to be challenging because there is still no consensual measurement applicable across all species. Up to now, a raw approximation consists in considering brain size as a proxy for cognitive abilities, with larger size considered equivalent to more advanced cognitive abilities [@Benson-Amram2016]. Yet, the brain is a mosaic of areas cognitively specialized [@barton2000mosaic]. These areas are likely to be under different selective pressures, and thus, to follow different evolutionary paths. In particular, they might be differently affected by species sympatry. First, brain areas involved in the storing of spatio-temporal information, such as the Hippocampus, home of an associative memory used for spatio-temporal navigation [@burgess2002human], should in particular be affected if the resource landscape is complexified by sympatric species foraging on the same resource (Hypothesis 1: memory is affected). Second, brain areas involved in processing more immediate sensory information, such as the Main Olfactory Bulb (MOB), Cerebellum [@koziol2014consensus; @sokolov2017cerebellum], and the Neocortex [@wiltgen2004new] should be particularly affected if the landscape of cues varies with sympatry (Hypothesis 2: cue processing is affected). Besides individual foraging, cognition can also be triggered by direct "social" interactions with other individuals [@byrne1994machiavellian; @dunbar2017there]. The Striatum is a brain area stimulated during social interactions [@baez2013role], and should, therefore, be affected by direct social interactions with heterospecifics (Hypothesis 3).

@@

|   Here, we investigated the intertwining between species sympatry and cognition using frugivorous primates as a study example. Frugivorous primates are an interesting group for such a question because fruits are the archetype of a hard-to-find resource yet predictable [@janmaat2016spatio], for which cognition thus considerably shapes the foraging strategy [@trapanese2019and]. To infer the effect of species sympatry on brain size evolution within frugivorous primates, we evaluated the support for evolutionary models accounting or not for sympatry, and investigated the directionality of the selection induced by sympatry history upon brain size evolution. Finally, we tested for correlative patterns between brain size or current sympatry and the evolutionary consequences (assumed to be reflected by species diversification) in all primates.

We gathered a large database on the whole brain size, the size of the MOB, Cerebellum, and the Neocortex, involved in immediate sensory information processing, and the Striatum and the Hippocampus, related to either social and foraging-related memory, containing between `r minAllAreas` to `r maxEQ` frugivorous primate species (depending on the brain area considered). After pondering by whole-body mass, we observed ample variations in brain area relative sizes. For instance, the lemuriformes, that are known to prioritize smell compared to other primate species, have the largest relative MOB size (Lemuriformes: mean $\pm$ SE = `r meanseMOBlemu[1]` $\pm$ `r meanseMOBlemu[2]`, other: `r meanseMOBother[1]` $\pm$ `r meanseMOBother[2]`, \@ref(fig:figbrain)). Similarly, platyrrhini, and callitrichine primates in particular, are known to form poly-specific associations [@heymann2000behavioural]. The latter show the highest relative size of the Striatum (Platyrrhini: mean $\pm$ SE = `r meanseStriatumplaty[1]` $\pm$ `r meanseStriatumplaty[2]`, other: `r meanseStriatumother[1]` $\pm$ `r meanseStriatumother[2]`, \@ref(fig:figbrain)).

|   To get the evolutionary history of species sympatry between frugivorous lineages, we first reconstructed primate biogeography history (N$_{species}$ = `r minRange`; [@matzke2013probabilistic; @matzke2016stochastic]) when considering `r length(areaName)` biogeographic areas (Figure \@ref(fig:figmap); [@kamilar2009environmental]) and their diet evolution (N$_{species}$ = `r minFruit + minLeaf` to `r maxFruit + maxLeaf`; [@bollback2006simmap]). We then calculated the likelihoods of models considering the role of species sympatry in the evolution of either the whole brain (using the encephalic quotient, EQ), or the size of the aforementioned specific brain areas relative to the whole-body mass (Figure. \@ref(fig:figbrain)). We specifically considered the matching competition (MC) model [@nuismer2015predicting] and density dependence models (DD$_{lin}$ and DD$_{exp}$), which assume a linear or exponential dependence of the brain sizes evolutionary rates on the number of sympatric lineages ([@drury2016estimating] see [Models of trait evolution: does species sympatry shape brain size evolution?]). We compared the support of models considering species sympatry to the support of simpler models assuming no effect of species sympatry, like the Brownian Motion (BM), the Ornstein-Uhlenbeck process (OU) considering that traits are constrained around an optimal value (e.g. stabilizing selection; see [@blomberg2020beyond] for a review), or the Early-Burst model (EB, [@blomberg2003testing]), the latter allowing to check for a time-dependence of the evolutionary rate, hence emphasizing that, if any, the density dependence is not an artefact due to time dependence. Support for each model was evaluated using an information-theoretic framework [@burnham2002model] based on the weights of the Akaike Information Criteria corrected for small samples (AICc) when considering all six models (MC, DD$_{lin}$, DD$_{exp}$, BM, OU, EB, see [Models of trait evolution: does species sympatry shape brain size evolution?]).

|   We found that models not considering species sympatry were the most likely in describing the evolutionary history of the EQ, the Neocortex, and the Cerebellum (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)), two areas specifically involved in immediate sensory information processing [@wiltgen2004new; @koziol2014consensus; @sokolov2017cerebellum], and also in memory consolidation for the Neocortex [@wiltgen2004new]. The fact that these biggest areas are best described by the Ornstein-Uhlenbeck process suggests a stabilization towards an optimal size, which may illustrate the trade-offs between costs and benefits of brain development [@isler2009expensive]. By contrast, density-dependence models considering sympatry were best supported in the foraging-related and social-related areas respectively: the Hippocampus, specialised in spatio-temporal memory [@burgess2002human] and the Striatum, involved in social interactions [@baez2013role]. The fact that we inferred positive rates *r* of density-dependence (Figure \@ref(fig:figresultsevolution)), suggested an acceleration of the evolutionary tempo of trait evolution together with increased diversity of sympatric lineages for the Hippocampus and the Striatum. The Main Olfactory Bulb (MOB), the area involved in sensory abilities, also tended to be best fitted by models considering sympatry as a whole: yet, Brownian-Motion was as likely as density-dependent or MC models, preventing firm conclusions on whether sympatry affected or not MOB size evolution (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)).

|   Next, to understand the effect of sympatry on brain sizes (i.e. toward “bigger” or “smaller” brains the more sympatric species), we fitted phylogenetic regressions (see [Phylogenetic regressions] a)). For these linear regressions, the predicted variable was the relative brain size values of the different areas. We considered as covariates the average percentage of overlapping range with other sympatric frugivorous species, as well as the number of such sympatric frugivorous species across their entire distribution range based on IUCN data [@IUCN]. On average ($\pm$ SE), the considered primate species had `r meanOverlap*100`% of their range overlapping with other species ($\pm$ `r seOverlap*100`). That ranged from `r minOverlapSpecies*100`% of overlap (*`r whichSpeciesMinOverlap`*), to `r maxOverlapSpecies*100`% of overlap (*`r whichSpeciesMaxOverlap`*). In terms of the distribution range, the considered primate species co-occurred on average with `r meanNspeciesCoocc` other primate species ($\pm$ `r seNspeciesCoocc`), ranging from `r minNspeciesCoocc` other species to `r maxNspeciesCoocc`. The number of sympatric species never influenced significantly the relative size of the brain or other specific areas (Table \@ref(tab:tableRegGradient)). Conversely, we found that the average percentage of overlapping range correlated with the relative size of areas that were better fit with models considering sympatry: the Hippocampus and the Striatum (Hippocampus: $t$`r testGradientHippocampus`; Striatum: $t$`r testGradientStriatum`). The correlations were all negative (Hippocampus: `r estimateGradientHippocampus`; Striatum: `r estimateGradientStriatum`), which means that higher range overlap between frugivorous species associates with lower relative size, insensitive to data and phylogenetic uncertainty (Table \@ref(tab:tabledfsensitivity), Figure \@ref(fig:forestPlot)). Given the acceleration of the evolutionary tempo with species sympatry ($r>0$ in the density-dependence models), it suggests that compared with isolated species, sympatric species are subject to a positive selection towards smaller brains, and not to a less intense selection for advanced cognitive abilities.

|   Finally, we investigated the evolutionary consequences of brain evolution and sympatry by evaluating whether brain sizes and sympatry level correlated with the net species diversification rates (defined as speciation minus extinction rates). To do so, we used lineage-specific birth-death models of species diversification [@maliet2019model; @maliet2020fast]. Overall, species diversification rates, estimated based on the primate molecular phylogeny, increased over time (Figure \@ref(fig:figdiversificationTime)), particularly in the early and late Miocene, around 25 and 11 Myr ago (Figure \@ref(fig:figdiversificationTime)). When accounting for phylogenetic dependence, no significant relationship between the net diversification rate and the relative size of brain areas was found (Table \@ref(tab:tableRegDiversification), Figure \@ref(fig:forestPlot); see robustness in Table \@ref(tab:tabledfsensitivity2)). Given the context-dependence of the direction of selection (towards bigger sizes when sympatry is low, smaller sizes otherwise), there is no surprise that we do not observe a correlation between the diversification rate and the three brain areas affected by species sympatry. Surprisingly however, we found no positive association between the net diversification rate and the EQ, nor with the relative sizes of the Cerebellum or the Neocortex, two areas insensitive to species sympatry too. This is puzzling because this contradicts a recent study [@melchionna2020macroevolutionary]. The visual inspection of the regressions however clearly evidenced a positive trend for the EQ and the Neocortex, if discarding phylogenetic non-independence (Figure \@ref(fig:figRegressionDiversification)). Indeed, sudden encephalisation in primates is clearly associated with a limited number of closely-related species [@decasien2017primate; @melchionna2020macroevolutionary]. This clearly limits the statistical power of our phylogenetically-corrected analyses, as we cannot decipher whether the connivance between brain size and species diversification results from a true biological link or appeared by chance. A positive association between brain size and diversification was also found in birds [@sayol2019larger] given that bigger brains act as a buffer to environmental challenges [@sol2007big]. This means that, despite what we found here, a positive association between brain size and species diversification evidenced in [@melchionna2020macroevolutionary] remains a likely possibility. Finally, although diversification was uncorrelated with brain size in frugivorous primates, it was influenced by the sympatry context. In particular, phylogenetic regressions highlighted a negative effect of the number of sympatric species on the diversification rate (`r textEstOneVar(as.numcharac(results.df_diversificationAndSympatry[4,2:4]))`, `r textTestOneVar(as.numcharac(results.df_diversificationAndSympatry[4,5:6]), statistics="t")`, Table \@ref(tab:tableRegDiversificationAndSympatry), Figure \@ref(fig:forestPlot)). In other words, the higher the number of sympatric species, the lower the diversification rate, a density-dependence trend that is frequently observed in many tetrapod clades [@condamine2019assessing].

Bigger brains are not necessarily better. The size of the brain is subject to a compromise between the energy it incurs, and the increase of fitness it allows. This is clearly emphasized by the fact that the biggest areas, the Cerebellum and the Neocortex, as well as the whole brain (EQ), were best described by the Ornstein-Uhlenbeck process. This evidence might suggest a stabilisation towards an optimal size resulting from an equilibrium between costs and benefits. As the brain is regionally specialised [@barton2000mosaic], different brain areas could be under different selective pressures. This is evidenced, here, by the differences of relative sizes across lineages in the MOB or Striatum in particular. We further show that sympatry is one factor that affects the selective regime under which some brain region evolves: although the brain as a whole was insensitive to species sympatry, the latter nonetheless induced a change in the relative size of the Hippocampus and the Striatum. These areas are involved in individual-based and social-based information processing, pinpointing that the two components might be under strong selection in primates [@decasien2017primate; @powell2017re; @gonzalez2018inference].

|   The negative influence of sympatry on the Striatum might be quite surprising. Direct interactions with heterospecifics should stimulate this area (Hypothesis 3). We observed that this area was relatively larger in callitrichines, particularly known to perform mixed-species groups [@heymann2015unlike]. When interacting with conspecifics, the Striatum underpins reward expectation, goal-directed behaviour, and planning abilities [@johnson2007integrating]. These skills are key within group, following a Machiavellian perspective [@byrne1994machiavellian]. Yet, these skills might also be relevant in a foraging context. It is thus possible that actions underpinned by the Striatum (e.g. planning/anticipating heterospecifics moves), might be key and up to now overlooked, when foraging in a multi-species context too.

|   Overall, the fact that only these two areas, particularly relevant to process and memorise spatio-temporal information, are sensitive to sympatry, is consistent with the idea of an effect of species affecting resource spatio-temporal patterns (Hypothesis 1). Competition is generally the first-thought mechanism to describe community structures [@rocha2015role] because it might affect the environment, and associate selective pressure, in which species evolve. We show that higher levels of sympatry is actually associated with smaller sizes of the Hippocampus or Striatum (in accordance with Prediction 1.2). This suggests that indirect competition for food might contribute to convoluting the environment which might generate a positive selection for smaller brains. Not only was brain evolution affected by sympatry, but sympatry induced a slowdown in diversification. Density-dependence within a clade is indeed generally associated with lower diversification rates [@moen2014does]. In particular, species competing for resources are thought to contribute to limiting their ranges [@price2009evolutionarily], hence constraining population size and their subsequent diversification [@pigot2013species]. These observations thus strengthen the idea of scramble competition between species that cascades both on species population dynamics and species cognition.

|   By contrast, potential indirect facilitation between species due to “social” cues (Hypothesis 2), is ruled out by the absence of an effect of sympatry on brain areas involved in immediate sensory information processing (e.g. Cerebellum or Neocortex). This absence of effect can stem from two possibilities. Either sympatry, indeed, does not affect the landscape of cues for foragers. Otherwise, it has been shown that foragers tend to use social information over environmental (i.e. personal) information, in particular in non-perfectly predictable environments [@rafacz2003environmental; @dunlap2016foraging]. Thus, if environmental complexity increases too much, “social” cues provided by heterospecifics might replace environmental ones. From this point of view, the size of the MOB, the Cerebellum, or the Neocortex should, at the very least, remain unaffected by sympatry. This nonetheless deserves further attention.

|   To conclude, the use of brain size as a proxy for cognition is a central debate with no optimal solution (see grounded criticism from [@deaner2000comparative; @healy2007critique; @logan2018beyond]). The current flourishment of consortia, allowing for much more detailed and standardized anatomical measurements (e.g. in primates: [@milham2018open]), or with standardized behaviourally explicit comparisons (e.g. on captive [@many2019establishing] or wild [@janmaat2021using] primates), might alleviate biases stemming from brain size analysis, but this will take time to generate large-enough datasets. In the meanwhile, brain size is a proxy much appreciated in practice, because of its easy accessibility for a “large” number of species. Here, we showed that species sympatry is an important factor shaping the evolutionary history of animals' brains, but the proximate mechanisms at play remain to be elucidated. Finally, it is very likely that any hypothesis on cognition evolution, generally discussed within species, could be broadened to a between-species context: foraging facilitation between species does exist [@olupot1998fruit; @Havmoller2021], and so do polyspecific social associations [@porter2001benefits], as well as inter-species territory defence [@drury2020competition; @losin2016ecological] or imitation and copying [@persson2018spontaneous; @pepperberg2002allospecific]. Similarly, prey-predator races could shape selection on cognitive abilities [@shultz2006chimpanzee]. As Alice said “It's a great huge game of chess that's being played—all over the world” ([@Carroll], Chapter II) and all individuals are just pieces to play with or against, no matter the species.

@@

Data processing, analyses, and plots were computed with R software (`r extractRversion()`, @Rsoftware). Used codes and data are freely available at [https://github.com/benjaminrobira/Meta_analysis_cognition_primates](https://github.com/benjaminrobira/Meta_analysis_cognition_primates). Note that in all these analyses, we discarded *Homo sapiens* and *Macaca sylvanus*.  The latter was discarded in the analyses because of geographical complete isolation and repeated intervention of human people in population maintenance [@modolo2005phylogeography]. A summary of available data per species is presented in Appendix Figure \@ref(fig:figmap).

## Data Collection

@@

We used a block of chronogram trees of the primate taxon of the 10kTrees project (downloaded on May 2021, version 3), as well as a consensus tree of 1000 trees for the subsequent phylogenetic analyses. The trees contain `r length(phylo_init$tip.label)` primate species.

@@

Brain data were obtained from [@decasien2019primate] for the  whole brain and all mentioned other regions (Cerebellum, Hippocampus, Main Olfactory Bulb (MOB), Neocortex, Striatum), [@powell2017re] and [@powell2019maternal] for the whole brain, Cerebellum and Neocortex size, [@todorov2019primate] for Hippocampus and Neocortex size, [@grueter2015home] for the whole brain size and [@navarrete2018primate] for the whole brain, Cerebellum, Hippocampus and Striatum size. They were freely available in the main manuscript or supplementary materials. When a species was represented multiple times within the dataset, we obtained a unique attribute by averaging it. From the global endocranial brain volume, we obtained the Encephalization Quotient (EQ, N$_{EQ,max}$ = `r maxEQ`) as follows [@decasien2017primate]

@@

with the brain volume in cm$^{3}$, 1.036 g/cm$^{3}$ being the assumed homogeneous brain density, and the body mass in g. EQ indicates whether the brain size ranges above (> 1) or below (< 1) expected given the body mass. Body mass was obtained from [@decasien2017primate; @powell2017re; @grueter2015home; @pearce2013space].
The sub-parts of the brain were chosen because they were involved in immediate sensory information processing (MOB, N$_{MOB,max}$ = `r maxMOB`), in movement and/or general information processing and retention (Neocortex, N$_{Neocortex,max}$ = `r maxNeocortex`, [@wiltgen2004new]; Cerebellum, N$_{Cerebellum,max}$ = `r maxCerebellum`, [@koziol2014consensus; @sokolov2017cerebellum]), short-term working memory and long-term spatio-temporal memory (Hippocampus, N$_{Hippocampus,max}$ = `r maxHippocampus`, [@burgess2002human]). The Striatum (N$_{Striatum,max}$ = `r maxStriatum`) supports information processing during social interaction, reward assessment, planning or goal-oriented behaviours [@baez2013role; @johnson2007integrating]. To investigate their evolutionary history, we used the ratio between their volume and body mass, so as to maximize comparability. As such, the use of specific region sizes relative to the body mass and not raw sizes depicts the evolution of cognitive abilities in terms of allocation rather than abilities per se (but see discussion in [@deaner2000comparative]). Percentage of frugivory and/or folivory was obtained based on a freely available dataset from [@decasien2017primate; @powell2017re; willems2013collective].

>>>>

#I follow: https://stackoverflow.com/questions/52918716/authors-and-affiliations-in-the-yaml-of-rmarkdown for author display
author:
#   - name: Benjamin Robira
#     email: benjamin.robira@normalesup.org
#     institute: [CEFE, MH]
#     correspondence: true
#   - name: Benoît Perez-Lamarque
#     email: benoit.perez@ens.psl.eu
#     institute: [ENS, MNHN]
# institute:
#   - CEFE: Centre d’Écologie Fonctionnelle et Évolutive, Université de Montpellier & CNRS, Montpellier, France.
#   - MH: Eco-anthropologie (EA), Muséum National d'Histoire Naturelle, CNRS, Université de Paris, Musée de l'Homme, Paris, France.
#   - ENS: Institut de Biologie de l’École Normale Supérieure (IBENS), École Normale Supérieure, CNRS, INSERM, Université PSL, Paris, France.
#   - MNHN: Institut de Systématique, Évolution, Biodiversité (ISYEB), Muséum National d’Histoire Naturelle, CNRS, Sorbonne Université, EPHE, Université des Antilles, Paris, France.

@@

output: 
 #      - '--lua-filter=scholarly-metadata.lua'
 #      - '--lua-filter=author-info-blocks.lua'
 bookdown::pdf_book:
 #bookdown::word_document2:
   number_sections: false
   toc: false
   citation_package: biblatex
   latex_engine: pdflatex
   fig_caption: true
   #pandoc_args:
     #- '--lua-filter=scholarly-metadata.lua'
     #- '--lua-filter=author-info-blocks.lua'
 keep_tex: true
always_allow_html: true

bibliography: bibliographyarticlepackage.bib
biblio-style: apa
biblatexoptions: [natbib=true]

@@

header-includes:
  - \usepackage{lettrine}
  # - \usepackage[nolists, nomarkers,tablesfirst]{endfloat} # For figures and tables at end
  - \usepackage{lineno} # For line numbering
  - \linenumbers # For line numbering
  - \usepackage{setspace}\doublespacing
  - \usepackage{fontawesome} #for fa symbols
  - \usepackage{tcolorbox}
  #- \pagenumbering{gobble} #for no page numbering
  - \pagenumbering{arabic} #for no page numbering
  - \setlength{\parskip}{0em} #to remove line gaps
  - \DeclareUnicodeCharacter{2212}{-}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \newcommand{\beginsupplement}{
      \setcounter{table}{0}  
      \renewcommand{\thetable}{S\arabic{table}} 
      \setcounter{figure}{0} 
      \renewcommand{\thefigure}{S\arabic{figure}}
    } 
  - \usepackage{titling}
  - \pretitle{
    \begin{flushleft}
    \rule[-0.15in]{0.25\linewidth}{0.8ex}
    \vspace{-0.8ex}
    \hrule
    \vspace{0.3in}
    \begin{LARGE}
    \noindent
    \textbf
    }
  - \posttitle{
    \end{LARGE}\newline
    \rule[-0.15in]{0.25\linewidth}{0.8ex}
    \hrule
    \end{flushleft}
    \vspace{0.2in}}
---

@@

```{r setup, include=FALSE, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(include=FALSE, echo=FALSE, message=FALSE)
knitr::opts_chunk$set(dpi=300) # Fig. resolution and size out.width = '100%', 
knitr::opts_chunk$set(fig.pos = 'p'#, fig.align = 'center'
                      ) # Places figures on pages separate from text, centered
knitr::opts_chunk$set(fig.env="figure") # Latex Fig. environment
knitr::opts_knit$set(eval.after = "fig.cap") #To insert R code into R Fig. caption
```

@@

```{r wordCount}
words <- RmdWords("Article_forsubmission.Rmd")
```

@@

**Abstract |**  
**Aim** The main hypotheses related to animal intelligence evolution highlight the role of conspecifics. Yet, space is often simultaneously occupied by species sharing the same ecological guild. These sympatric heterospecifics can compete for food, thereby stimulating or hampering cognition. Considering brain size as a proxy for cognition, we tested for the intertwining between species sympatry and cognition in primates.  
**Location** Central/South America, Africa, South/East Asia  
**Time period** From ~ 74 My ago to present  
**Major taxa studied** Primates  
**Methods** We compiled the size of several brain areas, diet and range data, and the phylogeny of more than 300 primate species based on published literature. We first retraced the evolutionary history of primate diet and biogeography. We then focused on frugivorous primates, and retraced the evolutionary history of the size of brain areas with evolutionary models considering or not sympatry.  
**Results** Models considering sympatry best predicted the evolution of brain areas related to long-term memory of interactions with the social or ecological environment, with a decrease of their size the higher the sympatry. By contrast, the whole brain or brain areas used in immediate information processing were best described by models not considering sympatry. Moreover, sympatry negatively affected primate diversification.  
**Main conclusions** This comparative study suggests that species sympatry contributes to shaping primate cognition and diversification. We speculate that this is due to an over-complexification of resource spatio-temporality in sympatry that counteracts the benefits of high cognitive abilities.

@@

<!-- Reviewer possibles: Decasien, Powell, Barton, Kamilar, Harmon, Drury -->
**Running title:** Species sympatry shapes primates' brain size  
\faTags \hspace{0.01in} **Keywords:** Brain size - Cognition - Competition - Co-occurrence - Diversification - Frugivory - Primates - Sympatry
<!-- \faInfoCircle  \hspace{0.01in} **Word Count:** `r words$num_words` \newline -->
<!-- Character Count: `r words$num_char` -->

@@

\initial{\textcolor{black}{C}}ognition evolution is shaped by the balance between socio-ecological drivers promoting cognitive abilities [@gonzalez2018inference] and physiological and energetic constraints limiting them [@navarrete2011energetics]. Primates are pivotal species for cognitive studies [@byrne2000evolution] because their cognition is thought to be promoted by interactions of individuals with conspecifics within the social unit [@byrne1994machiavellian; @dunbar2017there], among generations [@wilson1991molecular; @whiten2007evolution; @reader2002social; @herrmann2007humans; @tomasello2009cultural; @van2011social], between social units [@ashton2020interactions], or with the rest of their environment [@clutton1980primates; @milton1981distribution; @rosati2017foraging]. However, space is often occupied by many species belonging to the same ecological guild. Because of competition for food, we can predict that indirect interactions with sympatric heterospecifics are also likely to strongly shape the evolution of cognition, because they either affect the resource landscape by depleting it (Hypothesis 1), or the landscape of usable “social” cues to locate available food [@Avargues-Weber2013, Hypothesis 2]. In addition, direct interactions with heterospecifics, as much as with conspecifics [@dunbar2017there], might also affect cognition (Hypothesis 3).

|   Retracing the evolutionary history of cognitive abilities proves to be challenging because there is still no consensual measurement applicable across all species. Up to now, a raw approximation consists in considering brain size as a proxy for cognitive abilities, with larger size considered equivalent to more advanced cognitive abilities [@Benson-Amram2016]. Yet, the brain is a mosaic of areas cognitively specialized [@barton2000mosaic]. These areas are likely to be under different selective pressures, and thus, to follow different evolutionary paths. In particular, they might be differently affected by species sympatry. First, brain areas involved in the storing of spatio-temporal information, such as the Hippocampus, home of an associative memory used for spatio-temporal navigation [@burgess2002human], should in particular be affected if the resource landscape is complexified by sympatric species foraging on the same resource (Hypothesis 1: memory is affected). Second, brain areas involved in processing more immediate sensory information, such as the Main Olfactory Bulb (MOB), Cerebellum [@koziol2014consensus; @sokolov2017cerebellum], and the Neocortex [@wiltgen2004new] should be particularly affected if the landscape of cues varies with sympatry (Hypothesis 2: cue processing is affected). Besides individual foraging, cognition can also be triggered by direct "social" interactions with other individuals [@byrne1994machiavellian; @dunbar2017there]. The Striatum is a brain area stimulated during social interactions [@baez2013role] and should, therefore, be affected by direct social interactions with heterospecifics (Hypothesis 3).

@@

|   Here, we investigated the intertwining between species sympatry and cognition using frugivorous primates as a study example. Frugivorous primates are an interesting group for such a question because fruits are the archetype of a hard-to-find resource yet predictable [@janmaat2016spatio], for which cognition thus considerably shapes the foraging strategy [@trapanese2019and]. To infer the effect of species sympatry on brain size evolution within frugivorous primates, we evaluated the support for evolutionary models accounting or not for sympatry, and investigated the directionality of the selection induced by sympatry history upon brain size evolution. Finally, we investigated the evolutionary  of the intertwining between brain size and sympatry, by testing for correlative patterns between brain size or current sympatry and the species diversification in all primates.

@@

Data processing, analyses, and plots were computed with R software [`r extractRversion()`, @Rsoftware].

@@

We used a block of chronogram trees of the primate taxon of the 10kTrees project (downloaded on May 2021, version 3), as well as a consensus tree of 1000 trees for the subsequent phylogenetic analyses. The trees contain `r length(phylo_init$tip.label)` primate species (Fig. \@ref(fig:phylogeny)). Note that in all these analyses, we discarded *Homo sapiens* and *Macaca sylvanus*. The latter was discarded because of geographical complete isolation and repeated intervention of human people in population maintenance [@modolo2005phylogeography]. A summary of available data per species is presented in Appendix Fig. S3.

@@

Data were pooled from previous literature surveys (see Appendix [Data availability]). Brain data were obtained from @decasien2019primate for the  whole brain and all mentioned other regions (Cerebellum, Hippocampus, Main Olfactory Bulb (MOB), Neocortex, Striatum), @powell2017re and @powell2019maternal for the whole brain, Cerebellum and Neocortex size, @todorov2019primate for Hippocampus and Neocortex size, @grueter2015home for the whole brain size and @navarrete2018primate for the whole brain, Cerebellum, Hippocampus and Striatum size. They were freely available in the main manuscript or supplementary materials. Percentage of frugivory and/or folivory was obtained based on a freely available dataset from @decasien2017primate, @powell2017re, and @willems2013collective. Availability of trait and distribution range for the `r length(phylo_init$tip.label)` primate species represented in the primate phylogeny of the 10kTrees project is depicted in Appendix Fig. S3. From the global endocranial brain volume, we obtained the Encephalization Quotient (EQ, N$_{EQ,max}$ = `r maxEQ`) as follows [@decasien2017primate]

@@

with the brain volume in cm$^{3}$, 1.036 g/cm$^{3}$ being the assumed homogeneous brain density, and the body mass in g. EQ indicates whether the brain size ranges above (> 1) or below (< 1) expected given the body mass. Body mass was obtained from @decasien2017primate, @powell2017re, @grueter2015home, @pearce2013space.
The sub-parts of the brain were chosen because they were involved in immediate sensory information processing (MOB, N$_{MOB,max}$ = `r maxMOB`), in movement and/or general information processing and retention [Neocortex, N$_{Neocortex,max}$ = `r maxNeocortex`, @wiltgen2004new; Cerebellum, N$_{Cerebellum,max}$ = `r maxCerebellum`, @koziol2014consensus; @sokolov2017cerebellum], short-term working memory and long-term spatio-temporal memory [Hippocampus, N$_{Hippocampus,max}$ = `r maxHippocampus`, @burgess2002human]. The Striatum (N$_{Striatum,max}$ = `r maxStriatum`) supports information processing during social interaction, reward assessment, planning or goal-oriented behaviours [@baez2013role; @johnson2007integrating]. To investigate their evolutionary history, we used the ratio between their volume and body mass, so as to maximize comparability. As such, the use of specific region sizes relative to the body mass and not raw sizes depicts the evolution of cognitive abilities in terms of allocation rather than abilities per se [but see discussion in @deaner2000comparative].
